Efficient static analysis of Marlowe contracts

### 2020-05-21 papers/static_analysis2020/paper.tex
Address more reviews by @simonjohnthompson

State of the art libraries like SBV \cite{erkok2019sbv} allow straightforward
static analysis of arbitrary Haskell functions, with very few restrictions
on how those functions are implemented. However, static analysis can be very
computationally expensive, because it essentially needs to check every
possible execution of the contract.

@@

This paper contributes a number of approaches that can be used when
optimising static analysis, with examples extracted from a case-study where
we applied these approaches, and an overview of the techniques used
to test the correctness of the optimisations.

@@

modifications are more fundamental approaches that require considerable
changes to the structure of the implementation. We consider two main ideas:
\begin{itemize}
\item \textbf{Reducing the search space by using normalised execution paths relevant
to the property}. Instead of using the search space to model inputs, we
use it to model possible executions, and we ensure that we only
represent each possible group of equivalent executions once, i.e:
executions with the same normal form (or equivalence partitioning).
\item \textbf{Minimizing the representation of inputs and outputs}. We can benefit
from encoding inputs and outputs as concisely as possible, and discarding
information that we can infer in other ways.
\end{itemize}
In the following sections, we introduce the semantics of Marlowe as
a case study (Section~\ref{sec:Marlowe-intro}), we cover a
general approach to static analysis (Section~\ref{sec:Making-Marlowe-symbolic}),
we then explore both lightweight (Section~\ref{subsec:Lightweight-modifications})
and heavyweight (Section~\ref{subsec:Heavyweight-modifications})
optimisation techniques in more detail, and we illustrate them with
examples of how they apply to Marlowe static analysis. In Section~\ref{sec:Testing},
we briefly discuss some ways in which property based testing can be
used to address potential problems that may arise from the implementation
of the heavyweight optimisations described in the previous section.
In Section~\ref{sec:Measurements}, we present empirical results that
show the effects of heavyweight optimisations on the execution time
and memory usage of Marlowe's static analysis.

@@

Each transaction may include a list of inputs, a set of signatures,
a slot interval (in which the transaction can be executed), a set
of input UTxOs (incoming money
and tokens), and a set of outputs (outgoing money and tokens). For
a transaction to be valid in Marlowe, the transaction must have the
same effect for every slot within that slot range (be deterministic).
For example, if a transaction has a minimum slot number which is before
a timeout, and a maximum slot which after a timeout then the transaction
will fail with the error \texttt{AmbiguousSlotInterval}, since it would
not be deterministic.

@@

Even if we could represent mutually recursive datatypes, it would
not be feasible to directly translate the semantics, because a more
general limitation of SMT solvers is that they cannot, in general,
deal with arbitrarily sized computations. In other words, if termination
of a symbolic function is bounded by a parameter that can be arbitrarily
large, SMT solvers will often fail to terminate when trying to validate
a property, since doing so often requires a proof by induction. We
discuss how to address this problem in Section~\ref{subsec:Adapting-the-semantics}.

@@

As it turns out, we do not need to implement the semantics of the
\texttt{Close} construct at all, because it is impossible for the
\texttt{Close} construct to produce a failed payment
(we have also proven this\footnote{\url{https://github.com/input-output-hk/marlowe/blob/master/isabelle/CloseSafe.thy} using Isabelle \cite{nipkow2002isabelle})
(last visited on 04/04/2020)}: it only pays whatever there is available.
So we can safely remove it from the analysis.

@@

The solution to this increase in execution time was to pass as input, to
the static analysis implementation, a fixed list, with each element corresponding
to a transaction and consisting of a tuple with four symbolic integers:
\begin{enumerate}
\item An integer representing the minimum slot
\item An integer representing the maximum slot
\item An integer representing the \texttt{When} case whose input is being
included in the transaction, where zero represents the timeout branch
(and there being no inputs to the transaction).
\item An integer representing the amount of money or tokens (if the input
is a \texttt{Deposit}), or the number chosen (if the input is a \texttt{Choice})
\end{enumerate}
If we do not need a transaction, we just set all four numbers to $-1$.

@@

If we have two static analysis implementations, we have another opportunity
for testing. In our case, we have one efficient implementation that
is very different from the semantics and one inefficient implementation
that is much closer to the semantics, and we can compare the results of the
two implementations. We can do this as shown in Figure~\ref{fig:Property-based-testing-for-equivalence}.
We generate random contracts, we feed them to both static analysis
implementations, and we compare the results. If the results are the
same then test passes, if they are different then one of the implementations
is wrong.

@@

\noindent
Unfortunately, at the time of writing, we do not have a completely unoptimised
version of the static analysis that we can use to compare, because the semantics
of Marlowe have changed since we implemented that version.

>>>>

Marlowe's static analysis relies on SMT solvers, which are able to check whether a
set of constraints is satisfiable very efficiently. Thanks to state of the art
libraries like SBV \cite{erkok2019sbv}, we can describe those constraints in a
high-level language. In the case of SBV, we can directly write properties as
Haskell functions, with very few restrictions on how those functions are
implemented; SBV will automatically translate those functions into a language
that SMT solvers can understand (namely, SMTLib format \cite{barrett2010smt}).

Nevertheless, static analysis can become very computationally expensive if
implemented naïvely, because it needs to consider every possible value of the
property we want to check (in our case the main property we want to check
is whether a Marlowe contract will have enough money to make all the
payments described by the contract). Indeed, high-level abstractions
are often a tradeoff with efficiency.

@@

This paper contributes a number of approaches that can be used when
optimising static analysis, with examples extracted from a case-study where
we applied these approaches, an approach to ensure correctness
of the optimisations that combines the usage of property-based testing
and verification using automated theorem provers, and empirical data
that measures the effect of some of the optimisations on the implementation
of our case-study.

@@

modifications are more fundamental approaches that require considerable
changes to the structure of the implementation. But these optimisations can also
translate in important reductions to execution time and memory usage, as
shown by the experiments we report in Section~\ref{sec:Measurements}.
We consider two main ideas:
\begin{itemize}
\item \textbf{Reducing the search space by using normalised execution paths relevant
to the property}. Instead of using the search space to model inputs, we
use it to model possible executions, and we ensure that we only
represent each possible group of equivalent executions once, i.e:
executions with the same normal form (or equivalence partitioning).
\item \textbf{Minimizing the representation of inputs and outputs}. We can benefit
from encoding inputs and outputs as concisely as possible, and discarding
information that we can infer in other ways.
\end{itemize}
In the following sections, we introduce the semantics of Marlowe as
a case study (Section~\ref{sec:Marlowe-intro}), we cover a
general approach to static analysis (Section~\ref{sec:Making-Marlowe-symbolic}),
we then explore both lightweight (Section~\ref{subsec:Lightweight-modifications})
and heavyweight (Section~\ref{subsec:Heavyweight-modifications})
optimisation techniques in more detail, and we illustrate them with
examples of how they apply to Marlowe static analysis. In Section~\ref{sec:Testing},
we briefly discuss some ways in which property based testing can be
used to address potential problems that may arise from the implementation
of the heavyweight optimisations described in the previous section.
In Section~\ref{sec:Measurements}, we present empirical results that
show the effects of heavyweight optimisations on the execution time
and memory usage of Marlowe's static analysis.

@@

Each transaction may include a list of inputs, a set of signatures,
a slot interval\footnote{We use a slot interval because it is very difficult to
know in practice in which slot the transaction is going to be included,
and we need transactions to be deterministic, so we cannot use individual
slot numbers.}, a set of input UTxOs (incoming money and tokens), and a
set of outputs (outgoing money and tokens). For
a transaction to be valid in Marlowe, the transaction must have the
same effect for every slot within that slot range (be deterministic).
For example, if a transaction has a minimum slot number which is before
a timeout, and a maximum slot which after a timeout then the transaction
will fail with the error \texttt{AmbiguousSlotInterval}, since it would
not be deterministic.

@@

Even if we could represent mutually recursive datatypes, it would
not be feasible to directly translate the semantics, because a more
general limitation of SMT solvers is that they cannot, in general,
deal with arbitrarily sized computations. In other words, if termination
of a symbolic function is not bounded, or is only bounded by a symbolic
parameter, SMT solvers will often fail to terminate when trying to validate
a property, since doing so often requires a proof by induction. We
discuss how to address this problem in Section~\ref{subsec:Adapting-the-semantics}.

@@

As it turns out, we do not need to implement the semantics of the
\texttt{Close} construct at all, because it is impossible for the
\texttt{Close} construct to produce a failed payment
(we have also proven this\footnote{\url{https://github.com/input-output-hk/marlowe/blob/master/isabelle/CloseSafe.thy} (last visited on 04/04/2020)} using Isabelle \cite{nipkow2002isabelle}): it only pays whatever there is available.
So we can safely remove it from the analysis.

@@

The solution to this increase in execution time was to pass as input, to
the static analysis implementation, a fixed list, with each element corresponding
to a transaction and consisting of a tuple with four symbolic integers:
\begin{enumerate}
\item An integer representing the minimum slot
\item An integer representing the maximum slot
\item An integer representing the \texttt{When} case whose input is being
included in the transaction, where zero represents the timeout branch
(and there being no inputs to the transaction).
\item An integer representing the amount of money or tokens (if the input
is a \texttt{Deposit}), or the number chosen (if the input is a \texttt{Choice})
\end{enumerate}
If we need fewer transactions, we will just add dummy transactions with all four numbers
set to $-1$ at the end of the list.

@@

If we have two static analysis implementations, we have another opportunity
for testing. In our case, we have one efficient implementation that
is very different from the semantics and one inefficient implementation
that is much closer to the semantics, and we can compare the results of the
two implementations. We can do this as shown in Figure~\ref{fig:Property-based-testing-for-equivalence}.
We generate random contracts using a custom QuickCheck \cite{quickcheck} generator,
we feed them to both static analysis implementations, and we compare the results.
If the results are the same then test passes, if they are different then
one of the implementations is wrong.

@@

\noindent
Unfortunately, at the time of writing, we do not have a completely unoptimised
version of the static analysis that we can use to compare, because the semantics
of Marlowe have changed since we implemented that version. However, our impression
from manual testing is that the impact of lightweight optimisations was much more modest than the impact of heavyweight optimisations.
